Title
=====

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

#### The goal of this analysis is to answer the following 2 questions:
1. Across the United States, which types of events are most harmful with 
respect to population health?
2. Across the United States, which types of events have the greatest economic
consequences?

## Data Processing
#### Getting Data:  
 - [Download Link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)  
 - [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
 - **Date:** Dec 20th, 2014  
 - Extracted and stored in the *data* folder of the repository.  
 - Load the data into R

```{r}
library(plyr)
library(ggplot2)
library(gridExtra)
```
```{r, cache=TRUE}
data2 <- read.csv("./data/repdata-data-StormData.csv")
data <- data2
dim(data)
names(data)
```

There are a total of **902297** observations of **37** variables. For this
analysis I will only consider the variables related to Event Type, 
Fatalities, Injuries and Damage.
 
#### Cleaning Data:
First I will remove the unused variables to have a more readable console output:
```{r}
data <- subset(data, select=c("EVTYPE","FATALITIES","INJURIES",
                              "PROPDMG","PROPDMGEXP", "CROPDMG","CROPDMGEXP"))
summary(data)
str(data)
```
To answer the first question I will take a look to the columns **FATALITIES** 
and **INJURIES**. I will aggregate **EVTYPE** by these 2 variables and plot it.

```{r}
pop.healt <- ddply(data, .(EVTYPE),summarize,
                   TOTAL = sum(FATALITIES) + sum(INJURIES))
head(pop.healt)
dim(pop.healt)
```

I'm not interested in events with zero casualties so I will remove it from the
DataSet

```{r}
pop.healt <- subset(pop.healt, TOTAL != 0)
dim(pop.healt)
```
220 obs is still too much for a quick overview of the top events. Let's look at
the quantiles to see if we can narrow to the top events.

```{r}
quantile(pop.healt$TOTAL)
```
It seems like the 25% of the Top events has more than 44.25 casualties. Let's 
narrow it a little bit more...

```{r}
pop.healt <- subset(pop.healt, TOTAL > 44.25)
dim(pop.healt)
qplot(pop.healt$EVTYPE, pop.healt$TOTAL)
```

Lets look at the top event
```{r}
pop.healt[which.max(pop.healt$TOTAL),]
```

Now let's plot the Top 10 Events ordered by TOTAL
```{r}
pop.healt <- arrange(pop.healt, TOTAL, decreasing = TRUE)[1:10,]
# Re-arrange the levels is mandatory for ggplot to plot the bars in the same
# order as the dataframe
pop.healt$EVTYPE <- factor(pop.healt$EVTYPE,levels = pop.healt$EVTYPE)

# Custom ggplot theme
theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme_minimal()

ggplot(pop.healt) + geom_bar(aes(EVTYPE,TOTAL),stat = "identity") + 
  xlab("Event Type") + ylab("Number of people affected") +
  ggtitle("Numbers of people Affected in the U.S. by Weather Events from 1995 to 2011") +
  theme
```

That looks great. But I can't sum apples and bananas (FATALITIES and INJURIES)
so let's look a little bit into the two variables by separated and how they impact
in the number of people affected

```{r}
fatalities <- ddply(data, .(EVTYPE),summarize,
                   FATALITIES = sum(FATALITIES))
injuries <- ddply(data, .(EVTYPE),summarize,
                   INJURIES = sum(INJURIES))

fatalities <- arrange(fatalities, FATALITIES, decreasing = TRUE)[1:10,]
injuries <- arrange(injuries, INJURIES, decreasing = TRUE)[1:10,]

fatalities$EVTYPE <- factor(fatalities$EVTYPE,levels = fatalities$EVTYPE)
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

fatalities
injuries
```

Perfect! A little bit of Copy-Paste but it's fine so far. Now lets try to plot
it together

```{r}
injuries.plot <- ggplot(injuries) + geom_bar(aes(EVTYPE,INJURIES), stat = "identity")+
  ggtitle("Numbers of people injured\n in the U.S.\n by Weather Events\n from 1995 to 2011") +
  theme

fatalities.plot <- ggplot(fatalities) + geom_bar(aes(EVTYPE,FATALITIES), stat = "identity")+
  ggtitle("Numbers of fatalities\n in the U.S.\n by Weather Events\n from 1995 to 2011") +
  theme

grid.arrange(fatalities.plot, injuries.plot, ncol = 2)

```

Interesting discovery! Injuries are (aprox) ten times bigger than Fatalities...
so summing them is not valid (maybe normalize both variables and add them..??). 

Anyways, regarding the first question:  

#### Across the United States, which types of events are most harmful with respect to population health?  
It's safe to say that **Tornados** are the most harmful events.










