Weather Event Damage Analysis
=============================

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

#### The goal of this analysis is to answer the following 2 questions:
1. Across the United States, which types of events are most harmful with 
respect to population health?
2. Across the United States, which types of events have the greatest economic
consequences?

## Data Processing
#### Getting Data:  
 - [Download Link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)  
 - [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
 - **Date:** Dec 20th, 2014  

Needed libraries
```{r}
library(plyr)
library(ggplot2)
library(scales)
library(gridExtra)
```

Download and Extract the file:
```{r}
setwd("~/Repos/RepData_PeerAssessment2/")

if (!"repdata-data-StormData.csv.bz2" %in% dir("./data/")) {
  download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile = "data/repdata-data-StormData.csv.bz2")
  bunzip2("./data/repdata-data-StormData.csv.bz2", overwrite=TRUE, remove=FALSE)
}else{
  print("File already Downloaded and extracted.")
}

data <- read.csv("./data/repdata-data-StormData.csv")
dim(data)
names(data)
```

There are a total of `902297` observations of `37` variables. For this
analysis I will only consider the variables related to Event Type, 
Fatalities, Injuries and Damage.
 
#### Data Processing:
First I will remove the unused variables to have a more readable console output:
```{r}
data <- subset(data, select=c("EVTYPE","FATALITIES","INJURIES",
                              "PROPDMG","PROPDMGEXP", "CROPDMG","CROPDMGEXP"))
summary(data)
str(data)
```

In a first look we can say that there aren't `NA` but both `PROPDMGEXP` and `CROPDMGEXP`
have trash values that will need a further clean.

------------------

To answer the first question I will take a look to the columns `FATALITIES` 
and `INJURIES`. I will aggregate `EVTYPE` by these 2 variables and plot it.

```{r}
pop.healt <- ddply(data, .(EVTYPE),summarize,
                   TOTAL = sum(FATALITIES) + sum(INJURIES))
head(pop.healt)
dim(pop.healt)
```

I'm not interested in events with zero casualties so I will remove it from the
DataSet

```{r}
pop.healt <- subset(pop.healt, TOTAL != 0)
dim(pop.healt)
```
220 obs is still too much for a quick overview of the top events. Let's look at
the quantiles to see if we can narrow to the top events.

```{r}
quantile(pop.healt$TOTAL)
```
It seems like the 25% of the Top events has more than 44.25 casualties. Let's 
narrow it a little bit more to have a clear big picture of the top events...

```{r}
pop.healt <- subset(pop.healt, TOTAL > 44.25)
dim(pop.healt)
```

Let's look at the top event:
```{r}
pop.healt[which.max(pop.healt$TOTAL),]
```

Now let's plot the Top 10 Events ordered by `TOTAL`
```{r}
pop.healt <- arrange(pop.healt, TOTAL, decreasing = TRUE)[1:10,]
# Re-arrange the levels is mandatory for ggplot to plot the bars in the same
# order as the dataframe
pop.healt$EVTYPE <- factor(pop.healt$EVTYPE,levels = pop.healt$EVTYPE)

# Custom ggplot theme
theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme_minimal()

ggplot(pop.healt) + geom_bar(aes(EVTYPE,TOTAL),stat = "identity") + 
  xlab("Weather Events") + ylab("Number of people affected") +
  ggtitle("Numbers of people Affected in the U.S. by Weather Events from 1995 to 2011") +
  theme
```

That looks great. But I can't sum apples and bananas (`FATALITIES` and `INJURIES`)
so let's look a little bit into the two variables by separated and how they impact
in the number of people affected

```{r}
fatalities <- ddply(data, .(EVTYPE),summarize,
                   FATALITIES = sum(FATALITIES))
injuries <- ddply(data, .(EVTYPE),summarize,
                   INJURIES = sum(INJURIES))

fatalities <- arrange(fatalities, FATALITIES, decreasing = TRUE)[1:10,]
injuries <- arrange(injuries, INJURIES, decreasing = TRUE)[1:10,]

fatalities$EVTYPE <- factor(fatalities$EVTYPE,levels = fatalities$EVTYPE)
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

fatalities
injuries
```

Perfect! A little bit of Copy-Paste but it's fine so far. Now lets try to plot
it together

```{r}
injuries.plot <- ggplot(injuries) + geom_bar(aes(EVTYPE,INJURIES), stat = "identity")+
  ggtitle("Numbers of people injured\n in the U.S.\n by Weather Events\n from 1995 to 2011") +
  theme

fatalities.plot <- ggplot(fatalities) + geom_bar(aes(EVTYPE,FATALITIES), stat = "identity")+
  ggtitle("Numbers of fatalities\n in the U.S.\n by Weather Events\n from 1995 to 2011") +
  theme

grid.arrange(fatalities.plot, injuries.plot, ncol = 2)

```

------------------

Now let's dig a little bit about the economic consequences. We have 4 columns:  
 - `PROPDMG`  
 - `PROPDMGEXP`  
 - `CROPDMG`  
 - `CROPDMGEXP`
 
Let's try to unify the columns in just 1 numeric Value: `TOTALDMG`. Both 
`PROPDMGEXP` and `CROPDMGEXP` columns have a multiplier for each value in `PROPDMG` and
`CROPDMG`. We have Hundred (H), Thousand (K), Million (M) and Billion (B). We also
have trash values like -, +, ?

```{r}
data$PROPDMGEXP <- tolower(as.character(data$PROPDMGEXP))
data$CROPDMGEXP <- tolower(as.character(data$CROPDMGEXP))

data[data$CROPDMGEXP == "b", "CROPDMGEXP"] <- 9
data[data$CROPDMGEXP == "m", "CROPDMGEXP"] <- 6
data[data$CROPDMGEXP == "k", "CROPDMGEXP"] <- 3
data[data$CROPDMGEXP == "", "CROPDMGEXP"] <- 0
data[data$CROPDMGEXP == "?", "CROPDMGEXP"] <- 0
data$CROPDMGEXP <- as.numeric(data$CROPDMGEXP)
table(data$CROPDMGEXP)

data[data$PROPDMGEXP == "b", "PROPDMGEXP"] <- 9
data[data$PROPDMGEXP == "m", "PROPDMGEXP"] <- 6
data[data$PROPDMGEXP == "k", "PROPDMGEXP"] <- 3
data[data$PROPDMGEXP == "h", "PROPDMGEXP"] <- 2
data[data$PROPDMGEXP == "", "PROPDMGEXP"] <- 0
data[data$PROPDMGEXP == "?", "PROPDMGEXP"] <- 0
data[data$PROPDMGEXP == "-", "PROPDMGEXP"] <- 0
data[data$PROPDMGEXP == "+", "PROPDMGEXP"] <- 0
data$PROPDMGEXP <- as.numeric(data$PROPDMGEXP)
table(data$PROPDMGEXP)
```

Excellent! Now we have to calculate the `TOTALDMG`:  

```{r}
crop.total.dmg <- data$CROPDMG * (10 ^ data$CROPDMGEXP)
prop.total.dmg <- data$PROPDMG * (10 ^ data$PROPDMGEXP)

total.dmg <- crop.total.dmg + prop.total.dmg

data$TOTALDMG <- total.dmg
```

Now we just need to summarize the data and create a nice plot.

```{r}
event.dmg <- ddply(data, .(EVTYPE), summarize,
                   TOTALDMG = sum(TOTALDMG))
event.dmg <- arrange(event.dmg, TOTALDMG, decreasing = TRUE)[1:10,]

event.dmg$EVTYPE <- factor(event.dmg$EVTYPE, levels = event.dmg$EVTYPE)

ggplot(event.dmg) + geom_bar(aes(EVTYPE,TOTALDMG), stat="identity") +
  xlab("Weather Events") + 
  scale_y_continuous("Total Damage in Dollars",labels = comma) + 
  ggtitle("Total Damage in Dollars in the U.S,\n by Weather Events from 1995 to 2011") + 
  theme

```

## Results

#### Across the United States, which types of events are most harmful with respect to population health?  
It's safe to say that `TORNADOS` are the most harmful events, followed by 
`EXCESIVE HEAT`. 

#### Across the United States, which types of events have the greatest economic consequences? 
It's safe to say that `FLOODS` are the most harmful events, followed by 
`HURRICANE/TYPHOON`  










